---
title: "A Famous Math Problem"
author: "PSYC417"
format: html
editor: visual
---

## The Problem

A few years ago, a YouTube video became somewhat viral because a Taiwan educator gave their young students a seemingly simple but very challenging math problem. (One that many college students even struggled to solve.) The problem is this: ABCDE \* A = EEEEEE (this is six Es, it can be easy to miss). In other words, A, B, C, D, and E are all different digits, and when combined into a single number and multiplied by A, you get EEEEEE. 0 is not an acceptable digit. There is only one solution.

## Hints

You can solve this with for loops or while loops. You will either need to use nested for loops or a single while loop.

### for loop

-   You should have a loop for each letter (i.e., A, B, C, D, and E. Code should only be in the innermost loop.

### while loop

-   If you are using a while loop, you will need the following code:

```{r}
vec <- c(1:9)  
rsamp <- sample(vec,size=5)
a <- rsamp[1]
b <- rsamp[2]
c <- rsamp[3]
d <- rsamp[4]
e <- rsamp[5]
# This will draw a random number for each letter out of the numbers 1:9.
```

### Regardless of your approach, you will need:

-   Conditional statements.

-   The paste command (both for numbers, and for the output). Remember: You can use as.numeric and paste0 (or paste with the sep="" for numbers).

## Final Instructions

You should print the statement: "Eureka! I found the answer. ABCDE is \[number\], A is \[number\], and EEEEEE is \[number\]"

Happy programming!

```{r}

```
