---
title: "Lecture 2: Data Structures"
author: "Dr. Coutts"
format: html
editor: visual
---

## Data structures

-   Vectors: Unidimensional and can only contain a single data type. Created with the c() command.
-   Matrices: Two dimensional and can only contain a single data type. Created with the matrix() command.
-   Data frames: Two dimensional and can contain multiple data types. R's primary means of data storage and akin to a spreadsheet. Created with the data.frame() command.
-   Lists: Multidimensional and capable of storing other objects. Created with the list() command.

You can check the data structure if you're unsure about it.

```{r}
vec <- c("a","b","c","d")
mat <- matrix(c(1:9), nrow=3, ncol=3)
df <- data.frame(x = rep(TRUE, 4), y = c(1:4))
lis <- list(vec = vec, mat = mat, df = df)

is.vector(vec)
is.matrix(mat)
is.data.frame(df)
is.list(lis)
```

You can convert some objects to others using as.vector(), as.matrix(), etc.

```{r}
new_vec <- as.vector(mat);print(new_vec)
```

You can check lengths/dimensions using length() for vectors, lists, and matrices (produces number of elements), or nrow()/ncol() for matrices and data frames.

## Practice exercises

Create a different chunk for each of these practice exercises (for practice).

1.  Build each type of data structure. Try to vary the data types if you can. (Don't make every object consist of numbers and only numbers.)

2.  Convert each data structure to a different data structure. Print or view the new structures. Which ones DON'T work well?

3.  Double that numbers that exist in relevant vectors, matrices, or data frame columns. (Create ones with numbers if you must.)

4.  Add a new data frame column to your existing data frame that consists of boolean values. Name it "BOO."

5.  Store every object/structure you've created in a list called "complete." Subset the 3rd element of that list and assign it to an object called THIRD.

6.  Use the ratings and greetings vectors in the code chunk below to complete the following exercises:

    a.) Find how many "bad ratings" (a score of two or less) are in the ratings vector. Do this in one line of code.

    b.) Find how many "extreme ratings" there are in one line of code. (Hint: You may need to use an OR operator.)

    c.) Type greetings=="hey" and see what happens. Can you then find (in one line of code) whether more people say 'hey' than 'hi'? (Hint: This will require boolean operators!)

```{r}
ratings <- c(1,3,5,3,3,2,5,5,4,4,3,4,5)
greetings <- c("hey","hi","hey","yo","hello","sup","hey","hi")
```

**The end**
